---
title: "Homework 1"
author: "Theodore Sturiale"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, error = TRUE)
library(tidyverse)
```

# Preliminaries

- This file should be in `STAT240/homework/hw01` on your local computer.

While you should create your answers in the .Rmd file, homework problems are better formatted and therefore easier to read in the .html file. We recommended frequently knitting and switching between the two as you read and solve problems.


# Problem 1

Consider a vector `a` which has three numeric elements, a vector `b` which has two numeric elements, and a single number on its own, `c`.

For each of the following operations, write down how many elements long the output will be.

Think through the question first and make an initial guess, then check your work in the console. Also try to figure out which command will throw a warning message.

```{r}
a <- c(1, 2, 3)
b <- c(4, 5)
c <- 6
```

(a) `a + c`

```{r}
#The sum of a + c will be the vector a, with the value of c (6) added to each item within the vector. This will produce an output of 7, 8, 9
a + c
```


(b) `sum(a)`

```{r}
#Sum a will add together the items within the 'a' vector (1, 2, 3). These numbers sum to 6.
sum(a)
```


(c) `sum(a * c)`


```{r}
#sum(a*c) will multiply every item in the 'a' vector by the value in 'c' (6). Once the new vector of multiplied items has been computed, it will take the sum of the three items in the vector.
sum(a*c)
```


(d) `a + b`

```{r}
#When we sum 'a+b', the first value of vector 'a' will be added with the first item in vector 'b.' The same thing will happen for the second item. So the first two values of the combined vector will be 5, 7. For the third item in vector 'a', there is no corresponding third item in vector 'b', as a result, the third item will be added to the first item in vector 'b', giving us a final summed vector of 5,7,7. Since this addition isn't working as intended and has to loop around the b vector, this will throw a waring about the mismatched lengths.
a + b
```


(e) `sum(a) + b`

```{r}
#This operation first takes the sum of the vector 'a', which we have already computed to be 6. Then, it will add this value (6) to each of the items inside vector 'b'. This will produce a summed vector of 10, 11
sum(a) + b
```


# Problem 2

(a) Create a vector of all the whole numbers from 1 to 50 (including the endpoints), and **save** it in a new variable called `firstFifty`.

```{r}
firstFifty <- c(1:50)
firstFifty
```


(b) First compute the mean of `firstFifty`.

```{r}
meanFirstFifty <- mean(firstFifty)
meanFirstFifty
```


    Then, create a new vector which is the result of subtracting that mean from each value in `firstFifty`.
    
```{r}
fiftyMinusMean <- c(firstFifty - meanFirstFifty)
fiftyMinusMean
```


    Finally, take the sum of that resulting vector.
    
```{r}
sum(fiftyMinusMean)
```


    Let that sum be **printed as output**, do not save it as a variable.

    *Note: This result is not just a special case with these specific numbers, this process will return that sum for any vector!*
    



# Problem 3

Compute the **standard deviation** of `firstFifty` without using the `sd` command. Let this single number be printed as output.

```{r}
mean(firstFifty)
fiftyMinusMean
fiftyMinusMean^2
sum(fiftyMinusMean^2)
sum(fiftyMinusMean^2) / (length(firstFifty) -1)
sqrt(sum(fiftyMinusMean^2) / (length(firstFifty) -1))
```


To compute the standard deviation, execute the following steps in this order:

- Compute the mean of `firstFifty` and subtract it from every element in `firstFifty`, as in the previous question.

- Square every element in that vector.

- Take the sum of that new squared vector; this should be close to 10,400.

- Divide that number by (the length of the vector minus 1). You can obtain the length of firstFifty with `length(firstFifty)`. Don't forget parentheses & order of operations!

- Finally, take the square root (`sqrt()`) of the number you got from the previous step.

    The mathematical notation for this is: $$\sqrt{\frac{\sum_{i=1}^n{(x_i - \bar{x})^2}}{n-1}}$$ where $x$ is your vector of interest, in this case `firstFifty`, $n$ is the length of $x$, and $\bar{x}$ is the mean of $x$. 

    We encourage you to use `sd(firstFifty)` to check your work.
    
```{r}
sd(firstFifty)
```


# Problem 4

We have provided you below with a vector containing all of the prime numbers from 1 to 50.

```{r}
# We have written this code for you, do not edit it
primes <- c(2, 3, 5, 7, 11, 13, 17, 19, 23, 29, 31, 37, 41, 43, 47)
```

Create a vector of logical elements - that is, `TRUE`s and `FALSE`s; which indicates whether each of the elements in `firstFifty` is a prime number.

Let this vector be **printed as output**, do not save it as a variable.

```{r}
firstFifty %in% primes
```



# Problem 5

Below we demonstrate the base R function `substr()`.

```{r}
substr("abcde", 2, 4)
```

Use help page for `substr()` (which you can access with `?substr`) to answer the following questions.

(a) What does the `substr()` function do?

```{r}
#The substr() function extracts or replaces substrings within a character vector
```


(b) What does `substr()` do to a non-character input?  Answer this by reading the help file, then test out the behavior with an example in R.

```{r}
#It will turn the number into a string, returning a combined string of the characters corresponding to the boundaries set
substr(12345, 2, 4)
```

(c) Use the `substr()` function and the vector defined below to create a new vector of three-letter abbreviations for all twelve months of the year.  Let this be printed as output.

```{r}
months <- c("January", "February", "March", "April",
            "May", "June", "July", "August",
            "September", "October", "November", "December")
#Using substr() function to extract characters 1-3 from each variable in the months vector
substr(months, 1, 3)
```


# Problem 6

Create a dataframe using the `tibble()` command with five columns:

- The first column should be called `char`, and contain the 26 values of `letters`,
    
- The second column should be called `pos`, and contain the whole number values 1, 2, ..., 25, 26.
    
- The third column should be called `isVowel`, and should contain 26 logical values according to whether `char` in that row is one of "a", "e", "i", "o", or "u" or not. We won't consider "y" as a vowel for simplicity.
     
- The fourth column should be called `isPrime`, and should contain 26 logical values according to whether `pos` in that row is a prime number (or not).

- The fifth column should be called `vowelAndPrime`, and should contain 26 logical values; `TRUE` when that row has both a vowel and a prime number, and `FALSE` otherwise.

Save this dataframe in a new variable called `alphanumeric`.

```{r}
#Define Vowels
vowels <- c("a","e","i","o","u")
#Define Prime numbers 1-26
primes <- c(2, 3, 5, 7, 11, 13, 17, 19, 23)

#Create alphanumeric tibble
alphanumeric <- tibble(
  char = letters,
  pos = 1:26,
  isVowel = char %in% vowels,
  isPrime = pos %in% primes,
  vowelAndPrime = isVowel & isPrime
)

print(alphanumeric)
```


# Problem 7

Once you have created `alphanumeric` in Problem 6, uncomment (delete the #) the lines of code in the following chunk and run it.

```{r}
alphanumeric
print(alphanumeric, n = Inf)
```

* Notice that `print` with the `n = Inf` output is specifying to print ALL (`Inf` = infinity) rows of the data frame, whereas just calling `alphanumeric` on its own defaults to 10 rows being printed.

Looking at this output, Which vowel(s) are in a prime number position? In other words, in which rows is `vowelAndPrime` TRUE?

```{r}
#e is the only vowel in a prime number position
```

# Problem 8

The five lines of code below all return an error message.  Read the error messages, fix each line of code, and briefly describe how you resolved each error.

(a) We want to get the first element of the `primes' vector.

```{r}
#I switched the parenthesis for brackets to access the variable in the index
primes[1]
```

(b) We want to define a character vector with two elements, `apple` and `banana`.

```{r}
#I moved one of the parenthesis so that "apple" was enclosed in parenthesis before the comma
fruits <- c("apple", "banana")
```

(c) We want to set `notAPrimeNumber` equal to 51.

```{r}
#I replaces the == with <- to define the variable notAPrimeNumber as 51
notAPrimeNumber <- 51
```

(d) We want to get the first column of the built-in `mtcars` dataset.

```{r}
#I used brackets to get the first frist vector/column in the dataset
mtcars[1]
```

(e) We want to take a vector of ages in years and multiply each element by 12 to get a vector of ages in months.

```{r}
#Remove parenthesis from numbers so that they are not perceived as characters.
ages <- c(21, 8, 14, 35, 41)

ages * 12
```


# Submission

Once you have finished all the problems above, **knit** this file and submit **BOTH the .Rmd and .html files on Canvas!**

